%!TEX root = Beschreibung_des_Vorhabens.tex

%\todo{max. 10 publications}
\todo[inline]{In Bibtex umwandeln}
\todo[inline]{(2 Ko, 2 H, 6 gemeinsame, chronologisch rückwärts unabhängig vom Autor. Hier zu viele Beispiele, dies sind es aber noch nicht – gezielter suchen:}

\begin{itemize}
	\vspace{-1em} \item Experience and knowledge management in software engineering. K Schneider, Springer Science \& Business Media, 2009
	\vspace{-1em} \item Eliciting security requirements and tracing them to design: an integration of Common Criteria, heuristics, and UMLsec. SH Houmb, S Islam, E Knauss, J Jürjens, K Schneider, Requirements Engineering 15 (1), 63-93, 2009
	\vspace{-1em} \item Organizational learning and experience documentation in industrial software projects. D Landes, K Schneider, F Houdek, International journal of human-computer studies 51 (3), 643-661, 2009
	\vspace{-1em} \item Are developers complying with the process: an XP study. N Zazworka, K Stapel, E Knauss, F Shull, VR Basili, K Schneider, Proceedings of the 2010 ACM-IEEE International Symposium on Empirical, 1999
	\vspace{-1em} \item Effective experience repositories for software engineering. K Schneider, JP Von Hunnius, 25th International Conference on Software Engineering, 2003
	\vspace{-1em} \item Enhancing security requirements engineering by organizational learning. K Schneider, E Knauss, S Houmb, S Islam, J Jürjens, Requirements Engineering 17 (1), 35-562, 2003
	\vspace{-1em} \item  Feedback in context: Supporting the evolution of IT-ecosystems. K Schneider, S Meyer, M Peters, F Schliephacke, J Mörschbach, L Aguirre, International conference on product focused software process improvement
	\vspace{-1em} \item Sven Peldszus, Katja Tuma, Daniel Strüber, Jan Jürjens, Riccardo Scandariato: Secure Data-Flow Compliance Checks between Models and Code based on Automated Mappings. In: Proceedings of the 22nd ACM/IEEE International Conference on Model Driven Engineering Languages and Systems (MODELS), 2019.
	\vspace{-1em} \item Amir Shayan Ahmadian, Sven Peldszus, Qusai Ramadan, Jan Jürjens: Model-based Privacy and Security Analysis with CARiSMA. In: Proceedings of 2017 11th Joint Meeting of the European Software Engineering Conference and the ACM SIGSOFT Symposium on the Foundations of Software Engineering (FSE), pp. 989-993, September 2017.
\end{itemize}

\defbibheading{Pub}{\label{sec:pub}}
\vspace{-0.8cm}
\hspace{-0.95cm}
\begin{tabular}{p{8.3cm}p{8.3cm}}
	\defbibheading{PubKo}{\subsubsection*{Juerjens}}
	\addtopubju{Juerjens2018xyz}
	\printbibliography[heading=PubKo,category=publikationen_Juerjens] 
	&
	\defbibheading{PubSchn}{\subsubsection*{Schneider}}
	\addtopubschn{Schneider2013abc}
	\printbibliography[heading=PubSchn,category=publikationen_Schneider]
\end{tabular}\vspace{-0.2cm}